---
title: "gapminder_hw03"
output: github_document
editor_options: 
  chunk_output_type: console
---

# Homework 03: Use dplyr/ggplot2 to manipulate and explore data

Load packages

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

I selected some of the suggested tasks, as well as made up some of my own.

## Max/Min GDP per capita for all continents

Task 1: Get the maximum and minimum of GDP per capita for all continents.

```{r}
min_max <- gapminder %>%
  group_by(continent) %>%
  summarise(min = min(gdpPercap),
            max = max(gdpPercap))

min_max %>%
  knitr::kable()
```

Here are the minimum and maximum GDP per capita for all continents.

But I also want to find out what countries these belong to?

```{r}
min_max <- min_max %>%
  gather(key = min_or_max, value = "gdpPercap", min, max)

inner_join(min_max, gapminder, by = c("continent", "gdpPercap")) %>%
  select(1:4) %>%
  knitr::kable()
```

I used `gather` on the data and then `inner_join`. I can know see for each continent's min and max, what country that corresponds with.

```{r}
min_max %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_bar(aes(fill = min_or_max), stat="identity", position = position_dodge()) +
  scale_fill_manual(values = c("min" = "red", "max" = "darkgreen")) +
  theme_bw()
```

As a visual, we can see the min and max GDP per capita for each continent. The disparity is greatest in Asia annd lowest iin Oceania.

```{r}
min_max %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_bar(aes(fill = min_or_max), stat="identity", position = position_dodge()) +
  scale_fill_manual(values = c("min" = "red", "max" = "darkgreen")) +
  theme_bw() +
  scale_y_log10()
```

The same plot with a log y scale to better see the differences.

## Spread of GDP per capita

Task 2: Look at the spread of GDP per capita within the continents.

```{r}
gapminder %>%
  group_by(continent) %>%
  summarise(variance = var(gdpPercap),
            std_dev = sd(gdpPercap)) %>%
  knitr::kable()
```

Here, I've taken a look at spread by looking at variance and standard deviation. Like above, with the min/max, we can see the greatest variance in Asia. This highlights wealth inequality between the countries in Asia.

```{r}
gapminder %>%
  group_by(continent) %>%
  summarise(variance = var(gdpPercap),
            std_dev = sd(gdpPercap)) %>%
  ggplot(aes(x = reorder(continent, std_dev), y = std_dev)) +
  geom_bar(aes(fill = continent), stat="identity", show.legend = FALSE) +
  theme_bw()
```

Here is the standard deviation plotted as a bar plot. The x-axis is ordered by std_dev value. We can see that SD is lowest for African countries. However, does it make sense to look at SD? Are the values normally distributed?

```{r}
ggplot(gapminder, aes(x = continent, y = gdpPercap)) +
  scale_y_log10() +
  theme_bw() +
  geom_violin(aes(fill = continent), show.legend = FALSE) +
  geom_jitter(alpha = 0.2)
```

Here, I overlaid a fitter and violin plot. We can see that while the SD for African countries is low, there is a skew to high values. In fact Oceania has the tighest spread between its wealthiest and poorest countries.

## Life Expectancy over time in each continent

Task 3: How is life expectancy changing over time on different continents?

```{r}
gapminder %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_line(lwd = 0.5, show.legend = FALSE) + 
  facet_wrap(~ continent) +
  theme_bw()
```

Here, I plot the lifeExp of each country as a different colour of the years. Over time, life expenctancy increases, in general, for most countries and all continents. Additional things I did were to hide the legend, change the line width, change to bw theme, and facet by the different continents.

```{r}
gapminder %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_point() +
  facet_wrap(~ continent) +
  geom_smooth(method = "lm") +
  theme_bw()
```

To see trends for each continent, instead of lines for each country, I plotted things differently here. We have points for each country and a linear trend line added with `geom_smooth`. Now we can clearly see that for each continent, life expentancy increases over time.

## Mean vs Weighted mean Life Expectancy over time

Task 4: calculate the world-wide mean life expectancy (normal and weighted) over time

This is the mean life expectancy over time.

```{r}
mean_lifeExp <- gapminder %>%
  group_by(year) %>%
  summarise(mean = mean(lifeExp))

mean_lifeExp %>%
  knitr::kable()
```

This is the weighted mean by population.

```{r}
w_mean_lifeExp <- gapminder %>%
  group_by(year) %>%
  summarise(w_mean = weighted.mean(lifeExp, pop))

w_mean_lifeExp %>%
  knitr::kable()
```

We can compare the difference between the mean and weighted mean by population.

```{r}
inner_join(mean_lifeExp, w_mean_lifeExp, by = "year") %>%
  mutate(difference = mean - w_mean)%>%
  knitr::kable()
```

I used `inner_join` and `mutate` to do so. We can also see that generally, the mean is lower than the weighted mean.

We can plot both the mean and weighted mean over time.

```{r}
join_means <- inner_join(mean_lifeExp, w_mean_lifeExp, by = "year") %>%
  gather(key = mean_type, value = mean, mean, w_mean) %>%
  mutate(mean_type = if_else(mean_type == "w_mean", "weighted mean by pop", mean_type))

join_means %>%
  knitr::kable()

join_means %>%
  ggplot(aes(x = year, y = mean, colour = mean_type)) +
  geom_point() +
  geom_line() +
  theme_bw()
```

Here, I've made use of `inner_join` and `gather` to change the data frame. Then I used `mutate` and `if_else` (to rename "w_mean" to "weighted mean by pop"). I finally plot and see that both means increase over time (using geom_point and geom_line).

## Biggest drop and gain in GDP per capita

Task 5: Find out which country has biggest gain and biggest drop in GDP per capita over a 5 year period.


